---
title: "Analysis"
author: "Janette Avelar"
date: '2022-06-07'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(haven)
library(rio)
library(psych)
library(gmodels) #logistic regression functions
library(DT)
library(broom)
library(reshape2)
source(here("scripts", "functions.R"))

dat <- import(here("data", "clean_hsel_data.csv"))
```

Task 1: Descriptive

Present information on scores calculated for different components, including overall mean, SD, mins, and max.  
* instructional score max possible = 8   
* language support score max possible = 6    
* noninstructional service score max possible = 4   
* high school and parent service score max possible = 4  
* bilingual content score max possible = 6   
* overall resource max possible = 22   

```{r descriptive stats on scores}
options(scipen = 999)
max_scores <- describe(dat) %>% 
  select(mean, sd, min, max) %>% 
  rownames_to_column(., "var_desc") %>% 
  slice_tail(n = 12) %>% 
  mutate_at(vars(-c(1)), round, 2)
```


Task 1: Descriptive  
Understand how different services designed for ELs are distributed across region types within regions of the US.  
Rationale: Predict that the average number of services will vary.  

```{r regional services}
#prep data and get mean scores for each region
reg_desc <- dat %>% 
  select(region, ends_with(c("score"))) %>% 
  group_by(region) %>% 
  nest() %>% 
  mutate(avg_scores = list(pmap(data, ~mean(.x, na.rm = TRUE)))) %>% 
  select(-data)
#extract list elements
reg_avg <- data.frame(central = c(unlist(reg_desc$avg_scores[[3]])),
                southeast = c(unlist(reg_desc$avg_scores[[1]])),
                west = c(unlist(reg_desc$avg_scores[[2]])),
                northeast = c(unlist(reg_desc$avg_scores[[4]]))) %>% 
  #pretty labels for table
  rownames_to_column(., "Average Score") %>% 
  mutate(`Average Score` = str_replace_all(`Average Score`, "_", " "),
         `Average Score` = str_replace_all(`Average Score`, "instruct", "instructional"),
         `Average Score` = str_replace(`Average Score`, "hs", "high school and")) %>% 
  mutate_at(vars(-c(1)), round, 2)
#table
regional_service_scores_table <- datatable(reg_avg)
regional_service_scores_table
```

Now we'll create a second table to understand average number of services based on locale type.  

```{r locale services}
#prep data and get mean scores for each locale
locale_desc <- dat %>% 
  select(region_type, ends_with(c("score"))) %>% 
  group_by(region_type) %>% 
  nest() %>% 
  mutate(avg_scores = list(pmap(data, ~mean(.x, na.rm = TRUE)))) %>% 
  select(-data)
#extract list elements
locale_avg <- data.frame(suburb = c(unlist(reg_desc$avg_scores[[1]])),
                city = c(unlist(reg_desc$avg_scores[[2]])),
                town = c(unlist(reg_desc$avg_scores[[3]])),
                rural = c(unlist(reg_desc$avg_scores[[4]]))) %>% 
  #pretty labels for table
  rownames_to_column(., "Average Score") %>% 
  mutate(`Average Score` = str_replace_all(`Average Score`, "_", " "),
         `Average Score` = str_replace_all(`Average Score`, "instruct", "instructional"),
         `Average Score` = str_replace(`Average Score`, "hs", "high school and")) %>% 
  mutate_at(vars(-c(1)), round, 2)
#table
locale_service_scores_table <- datatable(locale_avg)
locale_service_scores_table
```

Task 2: Mixed Approach - Mean Difference

I'm interested in making a couple of visualizations (side-by-side bar charts most likely) that indicate the relationship between these average scores and the locale type, preferably within regions if I'm able to figure it out.    

```{r locale within region mean difference}
#mean diffs again, by double grouped nesting
mean_diffs <- dat %>% 
  select(region, region_type, ends_with(c("score"))) %>% 
  group_by(region, region_type) %>% 
  nest() %>% 
  mutate(avg_scores = list(pmap(data, ~mean(.x, na.rm = TRUE)))) %>% 
  select(-data)
#first bar chart should just show averages of each score nested within
#extract list elements
nested_avg <- data.frame(southeast_city = c(unlist(mean_diffs$avg_scores[[4]])),
                southeast_suburb = c(unlist(mean_diffs$avg_scores[[1]])),
                southeast_town = c(unlist(mean_diffs$avg_scores[[16]])),
                southeast_rural = c(unlist(mean_diffs$avg_scores[[13]])),
                west_city = c(unlist(mean_diffs$avg_scores[[2]])),
                west_suburb = c(unlist(mean_diffs$avg_scores[[14]])),
                west_town = c(unlist(mean_diffs$avg_scores[[5]])),
                west_rural = c(unlist(mean_diffs$avg_scores[[2]])),
                central_city = c(unlist(mean_diffs$avg_scores[[7]])),
                central_suburb = c(unlist(mean_diffs$avg_scores[[15]])),
                central_town = c(unlist(mean_diffs$avg_scores[[3]])),
                central_rural = c(unlist(mean_diffs$avg_scores[[10]])),
                northeast_city = c(unlist(mean_diffs$avg_scores[[8]])),
                northeast_suburb = c(unlist(mean_diffs$avg_scores[[11]])),
                northeast_town = c(unlist(mean_diffs$avg_scores[[9]])),
                northeast_rural = c(unlist(mean_diffs$avg_scores[[12]]))) %>% 
  rownames_to_column(., "Average Score") %>% 
  mutate(`Average Score` = str_replace_all(`Average Score`, "_", " "),
         `Average Score` = str_replace_all(`Average Score`, "instruct", "instructional"),
         `Average Score` = str_replace(`Average Score`, "hs", "high school and")) %>% 
  pivot_longer(cols = !`Average Score`,
               names_to = c("region", "locale_type"), 
               names_sep = "_",
               values_to = "score") %>% 
  pivot_wider(names_from = "locale_type",
              values_from = "score") %>% 
  unite("Average Score by Region", c(region, `Average Score`), sep = " ")

nested_region_locale_table <- nested_avg %>% 
  mutate_at(vars(-c(1)), round, 2) %>% 
  datatable()
nested_region_locale_table

#create loop for `nested_avg` object
# for (i in length(mean_diffs$avg_scores)){
#   
#   nested_avg <- rep(NA, length(mean_diffs$avg_scores*6))
#   
#   for(j in 1:length(nested_avg))
#   nested_avg[j] <- unlist(mean_diffs$avg_scores[i])
# }}

#create mean diffs across types
#collapse town/rural and suburb/urban for simplicity
nested_diffs <- nested_avg %>% 
  mutate(rural = (town + rural) / 2,
         urban = (suburb + city) / 2) %>% 
  select(-c(city, suburb, town)) %>% 
  #create percentage from max points for each score to make it meaningful
    #then create mean difference
#  instructional score max possible = 8   
#  language support score max possible = 6    
#  non instructional service score max possible = 4   
#  high school and parent service score max possible = 4  
#  bilingual content score max possible = 6   
#  overall resource max possible = 22 
  mutate(denom = c(rep(22, 4), rep(8, 4), rep(6, 4), rep(4, 4), rep(4, 4), rep(6, 4)),
         rural_pct = rural / denom,
         urban_pct = urban / denom,
         difference = raw_diff(urban_pct, rural_pct))

#prep data for plot
mean_diff_plot <- nested_diffs %>% 
  select(-denom) %>% 
  mutate(region = c(rep(c("southeast", "west", "central", "northeast"), 6))) %>% 
  rename(avg_score = `Average Score by Region`,
         pct_urban = urban_pct,
         pct_rural = rural_pct) %>% 
  mutate(avg_score = str_remove_all(avg_score, c("southeast ", "west ", "central ", "northeast "))) %>% 
  pivot_longer(cols = c(pct_rural, pct_urban),
               names_to = "locale",
               names_prefix = "pct_",
               values_to = "percent")
#labels for plot
diff_labs <- data.frame(y = mean_diff_plot$percent,
                        pct = c("", "7.83", "", "7.66", "", "11.53", "", "14.81", "", "7.46", "", "6.05", "", "14.05", "", "15.40", "", "6.41", "", "8.98", "", "7.90", "", "15.04", "", "5.67", "", "12.13", "", "2.69", "", "9.19", "", "12.88", "", "4.41", "", "20.78", "", "18.89", "", "8.52", "", "6.04", "", "14.17", "", "17.41"))
#plot
faceted_mean_diff <- ggplot(mean_diff_plot, aes(avg_score, percent, fill = locale)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_manual(values = custom_cols) +
    geom_text(aes(x = avg_score, y = percent + .025, label = diff_labs$pct)) +
    labs(x = "",
         y = "Score Percentages",
         title = "Mean Difference in Locale Scores by Region") +
    facet_wrap(vars(region)) +
    theme_final()
faceted_mean_diff
```

Pull out largest differences:

```{r largest diffs}
lrg_diff <- mean_diff_plot %>% 
  select(-c(rural, urban)) %>% #these columns keep confusing me...
  slice_max(difference, n = 10) %>% 
  unite("avg_score", c(avg_score, region), sep = " - ")
#almost all largest differences occur in the Northeast, that's interesting.
#additionally, parent and HS student services are represented twice, in NE and also in central regions.
#going to make a plot to indicate largest differences, with labels to mark these findings.
lrg_diff_labs <- data.frame(y = c(lrg_diff$percent),
                        labs = c("", lrg_diff$difference[1], "", lrg_diff$difference[3], "", lrg_diff$difference[5], "", lrg_diff$difference[7], "", lrg_diff$difference[9])) %>% 
  mutate(add = rep(.015, 10),
         y = y + add)

lrg_diff_plot <- lrg_diff %>% 
  ggplot(aes(avg_score, percent, fill = locale)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = custom_cols) +
  labs(x = "",
       y = "Score Percentages",
       title = "The Majority of the Largest Differences Occurred in the Northeast",
       subtitle = "additionally, high school and parent service scores appeared twice.") +
  geom_text(aes(x = avg_score, y = lrg_diff_labs$y, label = lrg_diff_labs$labs),
                color = "#c05750", size = 6) +
  theme_final()
lrg_diff_plot
```


Simple models  
Predict instructional score and language support scores (centered to get an idea of average) using region in the country as a predictor grouped within region types.  

```{r simple mods}
reg_desc_mod <- dat %>% 
  map_at(c(5:22), numeric_dichot) %>% 
  as.data.frame() %>% 
  group_by(region) %>%
  nest() %>% 
  mutate(instruct_model = map(data, ~lm(instruct_score_c ~ region_type, data = .x)),
         language_support_model = map(data, ~lm(language_support_score_c ~ region_type, data = .x)),
         resource_model = map(data, ~lm(resource_score_c ~ region_type, data = .x)),
         tidy_instruct = map(instruct_model, tidy),
         tidy_support = map(language_support_model, tidy),
         tidy_resource = map(resource_model, tidy),
         region_instruct_mean = map(data, ~mean(.x$instruct_score, na.rm = TRUE)),
         region_lang_support_mean = map(data, ~mean(.x$language_support_score, na.rm = TRUE))) %>% 
  ungroup()

#some kind of plot
resource_plot <- reg_desc_mod %>% 
  select(data) %>% 
  unnest(data) %>% 
  ungroup() %>% 
  select(region, ends_with(c("score_c"))) %>% 
  group_by(region) %>% 
  summarize_all(mean) %>% 
  melt(id = "region")

#labels to put in plot
resource_labs <- data.frame(x = rep("central", 24),
                            y = c(resource_plot$value),
                            lab = c("Overall Resources", rep("", 3), "Instructional Supports", rep("", 3), "Language Supports", rep("", 3), "Non-instructional Supports", rep("", 3), "Services for Parents/High Schoolers", rep("", 3), "Bilingual Content", rep("", 3))) %>% 
  mutate(add = c(rep(0, 4), 0.07, rep(0, 15), 0.12, rep(0, 3)),
         y = y + add)


ggplot(resource_plot, aes(region, value, color = variable, group = variable)) +
  geom_line(size = 1) +
  geom_text(aes(x = resource_labs$x, y = resource_labs$y, label = resource_labs$lab),
            position = position_nudge(x = -.15, y = -.03)) +
  scale_color_manual(values = custom_cols) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_final() +
  theme(legend.position = "none") +
  labs(x = "",
       y = "",
       title = "Differences in (Centered)* Scores by Region",
       caption = "*average indicated by dashed line")
```

Next to do is create tables from instruct/support/resource mods to show outcomes of modeling. Put asterisks for significant values.

[Ran a simple regression model to see if instructional supports, language-specific supports, or overall resource supports differed as a function of region. Significant results marked by asterisk.]

```{r model results}
#save models
#order = suburb, city, town, rural
reg_col <- c("Southeast", rep("", 3), "West", rep("", 3), "Central", rep("", 3), "Northeast", rep("", 3))
#create a for loop or something similar to build this argument for each group when you have more time
instruct_mods <- reg_desc_mod %>% 
  select(tidy_instruct) %>% 
  unnest(tidy_instruct) %>% 
  mutate(region = reg_col,
         term = str_remove(term, "region_type"),
         sig = ifelse(.05 > p.value & p.value > .01, paste0("*"), 
                      ifelse(.01 > p.value & p.value > .001, paste0("**"), 
                             ifelse(p.value < .001, paste0("***"), paste0("n.s."))))) %>% 
  mutate_at(vars(-c(region, term, sig)), round, 4) %>% 
  select(region, term, estimate, std.error, p.value, sig)

support_mods <- reg_desc_mod %>% 
  select(tidy_support) %>% 
  unnest(tidy_support) %>% 
  mutate(region = reg_col,
         term = str_remove(term, "region_type"),
         sig = ifelse(.05 > p.value & p.value > .01, paste0("*"), 
                      ifelse(.01 > p.value & p.value > .001, paste0("**"), 
                             ifelse(p.value < .001, paste0("***"), paste0("n.s."))))) %>% 
  mutate_at(vars(-c(region, term, sig)), round, 4) %>%
  select(region, term, estimate, std.error, p.value, sig)

resource_mods <- reg_desc_mod %>% 
  select(tidy_resource) %>% 
  unnest(tidy_resource) %>% 
  mutate(region = reg_col,
         term = str_remove(term, "region_type"),
         sig = ifelse(.05 > p.value & p.value > .01, paste0("*"), 
                      ifelse(.01 > p.value & p.value > .001, paste0("**"), 
                             ifelse(p.value < .001, paste0("***"), paste0("n.s."))))) %>% 
  mutate_at(vars(-c(region, term, sig)), round, 4) %>%
  select(region, term, estimate, std.error, p.value, sig)

#tables
instruct_mods_table <- datatable(instruct_mods)
support_mods_table <- datatable(support_mods)
resource_mods <- datatable(resource_mods)
```


Possible predictions:  
* could use logistic regression (or regular regression, but odds may be handier and you have binomial vars) to predict:  
+ urban areas will provide more resources  
+ schools with high concentrations of ELs in their district will provide more content instruction  
+ schools with high concentrations of ELs speaking other languages will provide more other language support

```{r regression models}
# regression model 1
# outcome = resource score (centered)
# predictor = number of students speaking majority non-English language
region_mod <- dat %>% 
  select(id, region_type, resource_score_c, receiving_content_primary) %>% 
  mutate(region_type = factor(region_type, levels = c("city", "suburb", "town", "rural"))) %>% 
  split(.$region_type) %>% 
  map(~lm(resource_score_c ~ receiving_content_primary, data = .)) %>% 
  map_dfc("coefficients") %>% 
  cbind(term = c("intercept", "slope_most", "slope_none", "slope_some"), .)
  #ggplot(aes(resource_score_c, receiving_content_primary)) +
  
# regression model 2
# outcome = resource score (centered)
# predictor = number of students speaking non-majority non-English language
# region_mod2 <- dat %>% 
#   select(id, region_type, resource_score_c, receiving_content_primary, receiving_content_other) %>% 
#   mutate(region_type = factor(region_type, levels = c("city", "suburb", "town", "rural"))) %>% 
#   split(.$region_type) %>% 
#   map(~lm(resource_score_c ~ receiving_content_primary + receiving_content_other, data = .)) %>% 
#   map_dfc("coefficients") %>% 
#   cbind(term = c("intercept", "slope_most", "slope_none", "slope_some"), .)

#compare mods
# anova(region_mod, region_mod2, test = "Chisq")

```

Maybe create a log model to consider differences in odds?