---
title: "Data Playground"
author: "Janette Avelar"
date: "3/10/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
library(here)
library(janitor)
library(haven)
library(rio)
source(here("scripts", "functions.R"))

raw_data  <- read_sas(here("data", "frss107puf.sas7bdat")) %>%
   janitor::clean_names()
# unknown <- read_sas(here("data", "formatspuf.sas7bdat")) %>% 
#   janitor::clean_names()
# unknown2 <- read_sas(here("data", "CntlinPUF.sas")) %>% 
#   janitor::clean_names()
```

## Final project requirements

Must include:  
* no code is used repetitively (2+) [10 points]  
* more than 1 variant of purrr::map used [5 points]  
* 1+ {purrr} function outside of basic `map` family [5 points]  
+ `walk_*`  
+ `reduce`  
+ `modify_*`  
+ etc.  
* at least 1 instance of parallel iteration [5 points]  
+ `map2_*`  
+ `pmap_*`  
* at least 1 use of `purrr::nest %>% mutate()` [5 points]  
* at least 2 custom functions [10 points]  
* code fully reproducible and housed on GH [10 points]  
* no obvious errors in output [5 points]  
* deployed on web and shareable via link [5 points]

## Potential variables of interest:  
###Demographics
`URBAN`  **could use this to create nested data**
+ 1 = city  
+ 2 = suburban  
+ 3 = town  
+ 4 = rural  
`OEREG`  **maybe**
+ 1 = Northeast  
+ 2 = Southeast  
+ 3 = Central  
+ 4 = West
`Q1` - Does your district currently enroll ELs at HS level?  
+ 1 = Yes  
+ 2 = No  **sort out 2s from the data - irrelevant**
`T_Q2` - What is the current total number of ELs enrolled in district?  
+ -8 = Inapplicable  
+ 1 = 1-10 ELs  
+ 2 = 11-100 ELs  
+ 3 = 101+ ELs

###Instructional language services
`Q3A` - Bilingual instruction in at least 1 content class?
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q3B` - Two-way bilingual/dual language programs in at least 1 content class?  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q3C` - ESL instruction in scheduled class periods?  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q3D` - ESL push-in or pull-out?
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q3E` - Instructional support by a para who *does not* speak native language?  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q3F` - Instructional support by a para who *does* speak native language?  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q3G` - Sheltered English/content instruction?  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q3H` - Other approaches  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  

###Additional language supports  
Programs or services designed specifically for ELs:
`Q11A` - Tutoring 
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q11B` - Summer school  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q11C` - Credit recovery course course/program  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q11D` - Mentoring  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q11E` - Distance education course/program  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q11F` - Other  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  

###Non-instructional language services
Services available for HS Els and their parents/guardians (most common native language):  
`Q12APT1` - written information about academic programs
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q12BPT1` - written information about CTE programs
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No 
`Q12CPT1` - translation services upon request for printed materials
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q12DPT1` - interpreters upon request for meetings or calls  
+ -8 = Inapplicable  
+ 1 = Yes  
+ 2 = No  
`Q13APT1` - apprx. ELs with most common native language (content instruction)  
+ -8 = Inapplicable  
+ 1 = No students  
+ 2 = Few students  
+ 3 = Some students  
+ 4 = Most or all students  
`Q13APT2` - apprx. ELs with most common native language (instructional support only)  
+ -8 = Inapplicable  
+ 1 = No students  
+ 2 = Few students  
+ 3 = Some students  
+ 4 = Most or all students 

Services available for HS Els and their parents/guardians (other native languages): 
`Q12APT2` - written information about academic programs  
+ -8 = Inapplicable  
+ 1 = All languages  
+ 2 = Some languages  
+ 3 = None  
`Q12BPT2` - written information about CTE programs  
+ -8 = Inapplicable  
+ 1 = All languages  
+ 2 = Some languages  
+ 3 = None  
`Q12CPT2` - translation services upon request for printed materials  
+ -8 = Inapplicable  
+ 1 = All languages  
+ 2 = Some languages  
+ 3 = None  
`Q12DPT2` - interpreters upon request for meetings or calls  
+ -8 = Inapplicable  
+ 1 = All languages  
+ 2 = Some languages  
+ 3 = None  
`Q13BPT1` - apprx. ELs with other languages (content instruction)  
+ -8 = Inapplicable  
+ 1 = No students  
+ 2 = Few students  
+ 3 = Some students  
+ 4 = Most or all students  
`Q13BPT2` - apprx. ELs with other languages (instructional support only)  
+ -8 = Inapplicable  
+ 1 = No students  
+ 2 = Few students  
+ 3 = Some students  
+ 4 = Most or all students  

Possible predictions:  
* could use logistic regression (or regular regression, but odds may be handier and you have binomial vars) to predict:  
+ urban areas will provide more resources  
+ schools with high concentrations of ELs in their district will provide more content instruction  
+ schools with high concentrations of ELs speaking other languages will provide more other language support

## Data cleaning  

Start by subsetting the data and saving a smaller workable datafile for analysis.

```{r subset}
sm_dat <- raw_data %>% 
  select(urban, oereg, q1, t_q2, #demographics
         q3a, q3b, q3c, q3d, q3e, q3f, q3g, q3h, #instructional language support
         q11a, q11b, q11c, q11d, q11e, q11f, #additional language support
         q12apt1, q12bpt1, q12cpt1, q12dpt1, q13apt1, q13apt2, #non-instructional primary
         q12apt2, q12bpt2, q12cpt2, q12dpt2, q13bpt1, q13bpt2) %>%  #non-instructional other
  filter(q1 == 1) #remove all HS that do not have EL services
#export(sm_dat, "hsel_data.csv")
```

Rename and recode cols

```{r rename}
dat <- import(here("data", "hsel_data.csv")) %>% 
  select(-q1)#drop forgotten col
#rename cols
new_col <- c("region_type", "region", "n_el", "b_bilingual_content", "b_twi_dual_content", "b_integrated_esl", "b_push_pull_esl", "b_mono_para", "b_bilingual_para", "b_sheltered_content", "b_other_instruct", "c_tutoring", "c_summer", "c_credit_recovery", "c_mentoring", "c_distance_ed", "c_other", "d_acad_info_primary", "d_cte_info_primary", "d_translation_primary", "d_interpreter_primary", "d_receiving_content_primary", "d_receiving_instruct_support_primary", "e_acad_info_other", "e_cte_info_other", "e_translation_other", "e_interpreter_other", "e_receiving_content_other", "e_receiving_instruct_support_other")
colnames(dat) <- new_col
#recode cols
dat <- dat %>% 
  mutate_all(na_if, -8) %>% 
  mutate(region_type = factor(region_type, levels = c(1, 2, 3, 4), labels = c("city", "suburb", "town", "rural")),
         region = factor(region, levels = c(1, 2, 3, 4), labels = c("northeast", "southeast", "central", "west")),
         n_el = factor(n_el, levels = c(1, 2, 3), labels = c("1-10", "11-100", "100+"))) %>% 
  mutate_at(vars(starts_with(c("b_", "c_", "d_")), -starts_with("d_receiving_")), factor, levels = c(1, 2), labels = c("yes", "no")) %>% 
  mutate_at(vars(starts_with(c("d_receiving", "e_receiving"))), factor, levels = c(1, 2, 3, 4), labels = c("no students", "few students", "some students", "most or all students")) %>% 
  mutate_at(vars(starts_with("e_"), -starts_with("e_receiving_")), factor, levels = c(1, 2, 3), labels = c("all languages", "some languages", "none")) %>% 
  #add ID number column for manipulation and matching
  rowid_to_column("id")
```

Create a composite `resources` score for outcome variable:  
* yes = 1  
* no = 0  
* all = 1  
* some = .5  
* none = 0  

Also create other subscores for subgroups:  
* `instruction_score` (b)  
* `language_support_score` (c)  
* `noninstruct_service_score` (d)  
* `hs_parent_service_score` (e)  
* all languages = 1  
* some languages = .5  
* none = 0  

```{r scores}
scores <- dat %>% 
  select(-c(2:4, 23, 24, 29, 30)) %>% 
  mutate_at(vars(-id), recode, "yes" = 1, "no" = 0, "all languages" = 1, "some languages" = 0.5, "none" = 0) %>% 
  mutate(resource_score = rowSums(.[setdiff(names(.),"id")], na.rm = TRUE)) %>% 
  rowwise() %>% 
  mutate(instruct_score = sum(c_across(starts_with("b_")), na.rm = TRUE),
         language_support_score = sum(c_across(starts_with("c_")), na.rm = TRUE),
         noninstruct_service_score = sum(c_across(starts_with("d_")), na.rm = TRUE),
         hs_parent_service_score = sum(c_across(starts_with("e_")), na.rm = TRUE)) %>% 
  select(id, ends_with("_score"))

centered_scores <- scores %>% 
  ungroup() %>% 
  mutate_at(vars(ends_with("_score")), center) %>% 
  rename_at(vars(-id), function(x) paste0(x, "_c"))
```

Final subscore:

`bilingual_content_score`
* bilingual_content = 1  
* twi_dual_content = 1  
* integrated_esl = 1  
* push_pull_esl = .5  
* mono_para = .5  
* bilingual_para = 1  
* sheltered_content = .5  
* other_instructional = .5  

*Note:* Score for instructional language services in this subgroup is based on the instruction score but gives a higher score to bilingual methods used. This is a composite measure made specifically for running a bilingual content instruction focused analysis because the generated score is based on prioritizing multilingual hiring and instruction, without penalizing services and supports solely in English.

```{r subscores}
bilingual_score <- dat %>% 
  select(id, starts_with("b_")) %>% 
  mutate_at(vars(c(2:4, 7)), recode, "yes" = 1, "no" = 0) %>% 
  mutate_at(vars(c(5, 6, 8, 9)), recode, "yes" = .5, "no" = 0) %>% 
  rowwise() %>% 
  mutate(bilingual_content_score = sum(c_across(starts_with("b_")), na.rm = TRUE)) %>% 
  ungroup() %>% 
  select(id, bilingual_content_score) %>% 
  mutate(bilingual_content_score_c = center(bilingual_content_score))
```

Join scores to dataframe and drop prefixes.

```{r join scores}
full_set <- reduce(list(dat, scores, centered_scores, bilingual_score), dplyr::left_join, by = "id") %>% 
  rename_at(vars(starts_with("b_")), function(x) str_remove(x, "^b_")) %>% 
  rename_at(vars(starts_with("c_")), function(x) str_remove(x, "^c_")) %>% 
  rename_at(vars(starts_with("d_")), function(x) str_remove(x, "^d_")) %>% 
  rename_at(vars(starts_with("e_")), function(x) str_remove(x, "^e_"))
#export
export(full_set, "data/clean_hsel_data.csv")
```

Create a data dictionary and export to data folder.
*in progress*

```{r dictionary}
#column names in final data set
renamed_cols <- colnames(full_set)
#column names in original data set
original_cols <- c("id", colnames(sm_dat)) %>% 
  str_remove("q1")
original_cols <-  unique(original_cols[original_cols != ""])
#labels for tables and plots
nice_labs <- renamed_cols %>% 
  str_replace_all("_", " ") %>% 
  str_replace_all("[:blank:]c$", " (centered)") %>% 
  str_replace_all("instruct", "instructional") %>% 
  str_replace_all("acad", "academic") %>% 
  str_replace_all("mono", "monolingual") %>% 
  map_chr(., toupper)
#original questions and descriptions tied to column
description <- import(here("data", "inc-data-info.csv")) %>% 
  clean_names %>% 
  rename(renamed_cols = variable_name)
map_chr(description$renamed_cols, tolower)
#bind all
dictionary <- list(renamed_cols, original_cols, nice_labs, description)


```

