---
title: "EL Services and Programs in the U.S."
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
    theme: readable
    social: menu
    source_code: https://github.com/jdavelar/653INC.git
editor_options: 
  chunk_output_type: console
---
```{r}
library(flexdashboard)
#library(shiny)
library(DT)
library(here)
source(here("scripts", "analysis.Rmd"))
```

Sidebar Title {.sidebar}
======================================================================
Author: Janette Avelar  
Course: EDLD 653 Functional Programming (Spring 2021)

*Project submitted to fulfill requirements for INCOMPLETE in Spring 2022.*

Data used was gathered from the Institute of Education Sciences (IES) through the National Center for Education Statistics (NCES).  

Data is publicly available through the NCES website as part of the "Programs and Services for High School English Learners" questionnaire implemented during the 2015-16 academic year. The survey was designed to be completed by the person or people at the district deemed most knowledgeable about programs and services for English Learner-classified students at the high school level.

Data Sources:
 
[Program and Services for High School English Learners Data](https://nces.ed.gov/pubsearch/pubsinfo.asp?pubid=2016150)

[NCES Full Report 2015-16](https://nces.ed.gov/pubs2016/2016150.pdf)

Project Purpose
=======================================================================
Column
-----------------------------------------------------------------------

### 

#### The State of English Learner Education in the U.S.

**What is the purpose of this project?**

The following visualizations are intended to.....

#### References

Viz 1 {data-orientation=rows}
=======================================================================

Column
-----------------------------------------------------------------------

### Descriptive Information

```{r}
regional_service_scores_table
```
